---
title: "WINN Tutorial: Drift and Batch Correction in LC-MS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WINN Tutorial: Drift and Batch Correction in LC-MS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This tutorial demonstrates **WINN** on simulated LC-MS data with known drift and batch effects. The goal is to show how WINN corrects technical artifacts **while preserving biological signal**.

We keep dependencies minimal and use only base R and the `winn` package.

## 1. Simulate LC-MS data

We simulate 50 metabolites measured across 200 samples, split into 4 batches. Every 8th measurement is a **QC sample** (pooled reference), so **all QC measurements should be essentially the same** in the baseline (true) signal. A subset of metabolites carries a true biological effect (group A vs B) in *non-QC* samples. We also simulate **sample‑wise dilution** (a realistic technical effect), then add drift, batch shifts, and noise.

```{r simulate}
set.seed(42)

n_met <- 50
n_samples <- 200
n_batches <- 4
samples_per_batch <- n_samples / n_batches

batch <- rep(seq_len(n_batches), each = samples_per_batch)
run_order <- seq_len(n_samples)

qc_idx <- seq(1, n_samples, by = 8)
non_qc_idx <- setdiff(seq_len(n_samples), qc_idx)

# Randomize biological groups along measurement order (QC excluded),
# with equal A/B counts **within each batch**
group <- rep(NA_character_, n_samples)
group[qc_idx] <- "QC"
for (b in seq_len(n_batches)) {
  idx_b <- which(batch == b)
  non_qc_b <- intersect(idx_b, non_qc_idx)
  n_b <- length(non_qc_b)
  labels <- rep(c("A", "B"), length.out = n_b)
  group[non_qc_b] <- sample(labels)
}

# True log-scale signal
baseline_log <- matrix(rnorm(n_met * n_samples, mean = 8, sd = 0.5), nrow = n_met)

# Biological effect on 10 metabolites (non-QC samples only).
# This is the *true* biological signal and is added before drift/batch/noise.
true_effect <- c(rep(0.8, 10), rep(0, n_met - 10))
true_log <- baseline_log
true_log[, group == "B"] <- true_log[, group == "B"] + true_effect

# QC samples: pooled reference (same baseline across QC samples)
pooled_log <- rowMeans(true_log[, group != "QC", drop = FALSE])
true_log[, qc_idx] <- pooled_log

# Add sample-wise dilution (multiplicative, technical artifact)
dilution_factor <- exp(rnorm(n_samples, mean = 0, sd = 0.12))

# Add drift per batch (realistic: warm-up → stable → mild decline, with non-linear curvature)
smoothstep <- function(x) {
  x <- pmin(pmax(x, 0), 1)
  x * x * (3 - 2 * x)
}

make_batch_drift <- function(n, shape = c("linearish", "nonlinear")) {
  shape <- match.arg(shape)
  # Piecewise structure: warm-up, plateau, decline
  n1 <- floor(n * 0.30)
  n2 <- floor(n * 0.45)
  n3 <- n - n1 - n2

  up <- smoothstep(seq(0, 1, length.out = n1))
  flat <- rep(1, n2)
  down <- 1 - smoothstep(seq(0, 1, length.out = n3))
  base <- c(up, flat, down)

  # Add non-linear curvature and small random walk (instrument instability)
  curve_strength <- ifelse(shape == "nonlinear", 0.15, 0.06)
  curve <- curve_strength * (seq_len(n) / n - 0.5)^2
  rw <- cumsum(rnorm(n, mean = 0, sd = ifelse(shape == "nonlinear", 0.012, 0.008)))
  base <- base + curve + rw

  # Re-center to keep drift around 0..1
  base <- base - min(base)
  base <- base / max(base)
  base
}

drift_log <- matrix(0, nrow = n_met, ncol = n_samples)
amp <- rnorm(n_met, mean = 0.18, sd = 0.06)
for (b in seq_len(n_batches)) {
  idx <- which(batch == b)
  shape <- ifelse(b %% 2 == 0, "nonlinear", "linearish")
  trend_b <- make_batch_drift(length(idx), shape = shape)
  drift_log[, idx] <- amp %*% t(trend_b)
}

# Add batch effects (additive shift + multiplicative scale on log scale)
batch_shift <- c(-0.4, 0.0, 0.3, 0.6)
batch_scale <- c(1.0, 1.1, 0.9, 1.2)

batch_log <- matrix(0, nrow = n_met, ncol = n_samples)
for (b in seq_len(n_batches)) {
  idx <- which(batch == b)
  batch_log[, idx] <- batch_shift[b] + log(batch_scale[b])
}

# Add noise
noise_log <- matrix(rnorm(n_met * n_samples, mean = 0, sd = 0.2), nrow = n_met)

# Observed data in intensity space
observed_log <- true_log +
  matrix(log(dilution_factor), nrow = n_met, ncol = n_samples, byrow = TRUE) +
  drift_log +
  batch_log +
  noise_log
true_intensity <- pmax(expm1(true_log), 0)
observed_intensity <- pmax(expm1(observed_log), 0)
```

## 2. Run WINN with fixed parameters, QC auto‑selection, and auto batch detection

We compare:

1. **Fixed parameters** (no QC guidance).
2. **QC-based auto parameter selection** using `control_samples`.
3. **Auto batch detection** (omit `batch`, let fkPELT segment the run).

```{r run-winn, message=TRUE}
library(winn)

winn_fixed <- winn(
  observed_intensity,
  batch = batch,
  run_order = run_order,
  median_adjustment = "shrink",
  remove_batch_effects = "anova",
  scale_by_batch = FALSE
)

winn_qc_auto <- winn(
  observed_intensity,
  batch = batch,
  run_order = run_order,
  control_samples = qc_idx,
  parameters = "auto",
  scale_by_batch = FALSE
)

winn_auto_batch <- winn(
  observed_intensity,
  batch = NULL,
  run_order = run_order,
  control_samples = qc_idx,
  parameters = "auto",
  median_adjustment = "shrink",
  remove_batch_effects = "anova",
  scale_by_batch = FALSE
)
```

## 3. Evaluate signal preservation

We compare corrected data to the **true signal** using:

- Mean per‑metabolite correlation with truth
- RMSE versus truth
- Preservation of group effect sizes (A vs B) on log1p scale (QC excluded)
- QC consistency (mean QC CV)

```{r metrics}
calc_metrics <- function(x, truth, group, qc_idx) {
  # Correlation per metabolite
  cor_vals <- sapply(seq_len(nrow(x)), function(i) {
    cor(x[i, ], truth[i, ], use = "pairwise.complete.obs")
  })
  mean_cor <- mean(cor_vals, na.rm = TRUE)

  # RMSE
  rmse <- sqrt(mean((x - truth)^2, na.rm = TRUE))

  # Effect size preservation on log1p scale
  x_log <- log1p(x)
  t_log <- log1p(truth)
  est_effect <- rowMeans(x_log[, group == "B", drop = FALSE]) -
    rowMeans(x_log[, group == "A", drop = FALSE])
  true_effect <- rowMeans(t_log[, group == "B", drop = FALSE]) -
    rowMeans(t_log[, group == "A", drop = FALSE])

  effect_cor <- cor(est_effect, true_effect, use = "pairwise.complete.obs")
  effect_rmse <- sqrt(mean((est_effect - true_effect)^2, na.rm = TRUE))

  # QC consistency metric (CV only)
  qc_log <- x_log[, qc_idx, drop = FALSE]
  qc_cvs <- apply(qc_log, 1, function(v) {
    mu <- mean(v, na.rm = TRUE)
    if (mu == 0) return(NA_real_)
    sd(v, na.rm = TRUE) / abs(mu)
  })
  qc_mean_cv <- mean(qc_cvs, na.rm = TRUE)

  data.frame(
    mean_cor = mean_cor,
    rmse = rmse,
    effect_cor = effect_cor,
    effect_rmse = effect_rmse,
    qc_mean_cv = qc_mean_cv
  )
}

results <- rbind(
  raw = calc_metrics(observed_intensity, true_intensity, group, qc_idx),
  winn_fixed = calc_metrics(winn_fixed, true_intensity, group, qc_idx),
  winn_qc_auto = calc_metrics(winn_qc_auto, true_intensity, group, qc_idx),
  winn_auto_batch = calc_metrics(winn_auto_batch, true_intensity, group, qc_idx)
)

results
```

## 4. Example: drift removal on 5 metabolites

Below are five example metabolites showing the **true signal**, the **observed (drifted/batched)** signal, and the **WINN‑corrected** signal (QC auto mode). We plot **log1p intensity** to reduce heteroscedasticity (common in LC‑MS) and add a smooth trend to highlight drift.

```{r plot, fig.width=8, fig.height=5}
library(ggplot2)

set.seed(7)
met_ids <- sort(sample(seq_len(n_met), 3))

df_plot <- data.frame(
  run_order = rep(run_order, times = 3 * length(met_ids)),
  metabolite = factor(rep(met_ids, each = n_samples * 3)),
  signal = rep(c("True", "Observed", "WINN (auto)"), each = n_samples, times = length(met_ids)),
  intensity = c(
    as.vector(t(true_intensity[met_ids, ])),
    as.vector(t(observed_intensity[met_ids, ])),
    as.vector(t(winn_auto_batch[met_ids, ]))
  )
)

df_plot$log1p_intensity <- log1p(df_plot$intensity)

ggplot(df_plot, aes(x = run_order, y = log1p_intensity, color = signal)) +
  geom_point(size = 0.7, alpha = 0.5) +
  geom_smooth(se = FALSE, method = "loess", span = 0.25, linewidth = 0.7) +
  facet_wrap(~ metabolite + signal, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c("True" = "#2c7fb8", "Observed" = "#636363", "WINN (auto)" = "#d95f0e")) +
  labs(
    title = "Three Example Metabolites: True vs Observed vs WINN-corrected",
    x = "Run order",
    y = "log1p(Intensity)",
    color = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## 5. Takeaways

- **WINN improves recovery of the true signal** compared to raw measurements.
- **QC-based auto selection typically improves QC consistency** while maintaining biological effect sizes.
- Providing known batch information generally performs better than auto‑detection.

This simulated dataset is intentionally simple. For real LC‑MS data, consider using QC samples and metadata to guide drift and batch correction.
